Recruiting Chapter

### 2020-10-05 manuscript.Rmd
draft changes front end manuscript

When we contrast research about enrollment management to descriptions of enrollment management practice, from investigative journalists or from consulting firms [cite], it becomes apparent that the research community is decades behind practice and has ignored a great number of enrollment management practices that affect access to higher education. Thus, the enrollment management remains an opaque industry to researchers and, in turn, to policymakers and the public. Developing thoughtful state and federal policies about enrollment management depends on researchers doing the slow, methodical work of developing an empirical literature that documents enrollment management practices and evaluates their effect on opportunities for students.

@@

Table X shows the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities. ^[table drawn directly from @RN4664 figure 9; based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions]. Institutions identify undergradaute prospects largely by purchasing "student lists" from College Board, ACT, and other vendors.  Table X indicates that private and public institutions spent 14% and 12%, respectively, of their undergraduate recruitment budget on student list purchases. Institutions utilize digital advertising, traditional advertising, and social media to solicit inquiries and to create positive "buzz" amongst potential stealth applicants [@RN4411][CITE OTHER]. <!-- Table X indicates that...  [summarize expenditure on stuff related to soliciting inquiries once you figure out answers to below questions] -->  Once identified, prospects/inquiries are targeted with face-to-face and remote recruiting interventions designed to solit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, email, and text messages. <!-- Table X indicates that these "prospective student communications" account for 13% of the recruiting budget at private institutions and 17% of the budget at public institutions.  -->Face-to-face recruiting interventions include on-campus visits (12% and 11% of budget, respectively, at private and public institutions) and off-campus recruiting events, often referred to as "travel" (17% and 16% of budget, respectively, at private and public institutions).
<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

@@

To date, our analyses have focused on recruiting visits by public research universities. As described below, the majority of public universities in our sample made far more out-of-state visits than in-state visits. These out-of-state visits focused on wealthy, predominantly white communities, with a disproportionate number of visits to private high schools. This chapter analyzes visits to private high schools. Using social network methods, we analyze visits by private colleges and universities to private high schools, visits by public research universities to private high schools, and then compare similarities and differences between public and private postsecondary institutions [as a means of saying something about "privatization"?]

@@

PARAGARPH SAYING HARD TO GET DATA ON RECRUITING BEHAVIOR

Empirical research depends on data but obtaining data about marketing and recruiting behaviors is difficult, at least through conventional data collection methods such as surveys and interviews.

PARAGRAPH SAYING WHAT WE DID TO OBTAIN DATA ON OFF-CAMPUS RECRUITNG

@RN5659 found that this is true. Here's an example of parenthetical citation [@RN4552].

>>>>

We present this chapter as one example of empirical research about enrollment management behavior. When we contrast research about enrollment management to descriptions of enrollment management practice, from investigative journalists or from consulting firms [cite], it becomes apparent that the research community is decades behind practice and has ignored a great number of enrollment management practices that affect access to higher education [AN EXCEPTION TO THE CRITIQUE IS A SMALL BUT HIGH-QUALITY CASE STUDY LITERATURE FROM SOCIOLOGY]. Thus, the enrollment management remains an opaque industry to researchers and, in turn, to policymakers and the public. Developing thoughtful state and federal policies about enrollment management depends on researchers doing the slow, methodical work of developing an empirical literature that documents enrollment management practices and evaluates their effect on opportunities for students. 
[TONE DOWN THE CRITIQUE; DON'T WANT TO MAKE IT SEEM LIKE WE THINK WE ARE THE FIRST TO STUDY THIS STUFF; GIVE CREDIT TO THE EXCELLENT CASE STUDY RESEARCH]

@@

[REVISE BELOW PARAGRAPH TO TAKE OUT MOST OF THE NUMBERS OF PERCENT SPENT ON EACH ACTIVITY; BOGS DOWN THE PROSE; MAIN POINT IS HIGHLIGHTING THE ACTIVITIES THAT TARGET SPECIFIC ELEMENTS OF THE FUNNEL; READER CAN READ THE PERCENTAGES ON THEIR OWN]
Table X shows the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities. ^[table drawn directly from @RN4664 figure 9; based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions]. Institutions identify undergradaute prospects largely by purchasing "student lists" from College Board, ACT, and other vendors.  Table X indicates that private and public institutions spent 14% and 12%, respectively, of their undergraduate recruitment budget on student list purchases. Institutions utilize digital advertising, traditional advertising, and social media to solicit inquiries and to create positive "buzz" amongst potential stealth applicants [@RN4411][CITE OTHER]. <!-- Table X indicates that...  [summarize expenditure on stuff related to soliciting inquiries once you figure out answers to below questions] -->  Once identified, prospects/inquiries are targeted with face-to-face and remote recruiting interventions designed to solit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, email, and text messages. <!-- Table X indicates that these "prospective student communications" account for 13% of the recruiting budget at private institutions and 17% of the budget at public institutions.  -->Face-to-face recruiting interventions include on-campus visits (12% and 11% of budget, respectively, at private and public institutions) and off-campus recruiting events, often referred to as "travel" (17% and 16% of budget, respectively, at private and public institutions).
<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

@@

To date, our analyses have focused on recruiting visits by public research universities. As described below, the majority of public universities in our sample made far more out-of-state visits than in-state visits. These out-of-state visits focused on wealthy, predominantly white communities, with a disproportionate number of visits to private high schools. This chapter analyzes visits to private high schools. Using social network methods, we analyze visits by private colleges and universities to private high schools, visits by public research universities to private high schools, and then compare similarities and differences between public and private postsecondary institutions [as a means of saying something about "privatization"?]
[MOTIVATE PARTICULAR FOCUS OF THIS CHAPTER: NETWORK ANALYSIS OF VISITS TO PRIVATE HIGH SCHOOLS BY BOTH PUBLIC AND PRIVATE COLLEGES/UNIVERSITIES]

@@

Our research analyzes off-campus recruiting visits by college and university admissions staff. So what do we know about off-campus recruiting? As is true for most aspects of enrollment management, much of what we know about off-campus recruiting comes from consulting firms (e.g., Ruffalo-Noel Levitz, EAB), professional associations (e.g., NACAC), and from practioner-oriented publications. Market research conceives of off-campus recruiting visits as a means of identifying prospects, deepening engagement with prospects already being targeted through mail/email, and maintaining relationships with guidance counselors at “feeder schools” [@RN4323; @RN4402; @RN4664]. With respect to expenditure, @RN4664 found that private (non-profit) 4-yr institutions spent an average of 17% of their undergraduate marketing/recruiting budget on "travel" to high schools and college fairs (as shown in Table X), a higher percentage of budget allocation than any other marketing/recruiting activity. Public institutions spent an average of 16% of their budget on travel, second only to "prospective student communications" at 17%. An emergent trend over the past decade -- partially a response to public universities seeking nonresident students -- has been the growth of "regional recruiters" who target specific metropolitan areas in the US and abroad. These regional recruiters may be college/university employees or they may be independent contractors who live in the metropolitan area they recruit.

@RN4402 documents the self-reported efficacy of marketing/recruiting interventions. For the median private 4-yr institution, off-campus visits were the second highest source of inquiries (after student list purchases), accounting for 17% of inquiries. Off-campus visits were tied with student list purchases as the highest source of enrollees, accounting for 18% of enrollees. For the median public institution, off-campus visits accounted for 19% of inquiries (second only to student list purchases) and accounted for 16% of enrollees (ranked third after stealth applicants and on-campus visits).

A sociological case-study literature develops helpful insights about the mechanisms and functions of off-campus recruiting visits. @RN4324 analyzes visits from the perspective of students at two racially and socioeconomically diverse public high schools. @RN4324 found that high school visits -- including college fairs, instant decision events, and small-group representative visits -- influenced where students applied and where they enrolled. This finding was strongest for first-generation students and under-represented students of color, who often reported that "school counselors had low expectations for them and were too quick to suggest that they attend community college." This trust vacuum created an opportunity for colleges because these students were drawn to colleges that connected with them and "made them feel wanted."  @RN4324 describes a high-achieving, first-generation, African American student who was admitted by a highly selective liberal arts college but chose to attend a less selective college that "seemed to want him more." By contrast, affluent students with college educated parents were less taken by such overtures and more concerned with college prestige.

Of particular importance for this chapter, @RN3519 and @RN4321;@RN4407 highlight the relational function of visits.  @RN3519 provides an ethnography of the admissions office at a selective private liberal arts college, highlights the relational function of visits. During the autumn "travel season," admissions officers visited selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54) because "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide" {@RN3519} [p. 54]. The College tended to visit the same schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and had the resources and motivation to host a successful visit. Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their affect on high school guidance counselors. The logic is that a guidance counselor who views a college favorably will steer students to the college.

@RN4321 analyzed recruiting from the perspective of an elite private boarding school in order to understand "how such schools continue to get comparatively under-qualified students into top colleges and universities." The answer to this question begins by thinking about the goals of colleges, which are represented by admissions officers, and the goals of private high schools, which are represented by guidance counselors. Colleges want high achieving students who can pay tuition and donate. Additionally, drawing from @@RN3519, colleges want a class composed of "interesting characters" whose curricular and extracurricular strengths meet the needs of different campus constituents (e.g., academic majors, the athletic department, clubs, etc.). Colleges also want low acceptance and high yield rates to move up the rankings. Private high schools want to send _all_ students to the best college possible. Here, high school guidance counselors face "the pressure of making sure their school seems worth it – that...paying some $40,000+, really does aid students in the college process" [FROM CHAPTER. FIND PG NUMBER]  The challenge counselors face is "some of these students are slightly better than others. These students will likely get into more than one school -- but they can only attend one. And this will lower the chances of your "second-best" students getting into top schools" [@RN4407] [p. 173-174].

"Luckily," @RN4321 [pg. xx] writes, "the problem for elite boarding schools matches up quite nicely with the problem faced by elite colleges." Specifically, college admissions officers receive applications from many outstanding students. But "These outstanding students will also be outstanding to Princeton, Yale, Stanford, and everywhere else. How do you know the ones you pick will attend your school? you can't quite trust applicants, as they are all likely to tell you how much they want to go to your school. And if students you accept go somewhere else, there's not much you can do. But you can get better information -- information you want -- from their high school" [@RN4407] [pg. 173?].

@RN4407 argues that the desire by colleges for trustworthy information about applicant intentions creates an opportunity for high school counselors to advocate on behalf of their students. This opportunity depends on guidance counselors having personal relationships with university admissions offices and on having small enough caseloads to advocate for each student individually. To explain how the process works, @RN4407 describes two hypothetical students -- Susan and Billy -- who both apply to Harvard and Yale.  Susan is a shoo-in at both universities, but wants to attend Harvard. Billy has a weaker academic record than most Ivy League students, but has strong extracurricular activities. The guidance counselor tells Harvard that Susan wants to attend Harvard. Next, he informs Yale that Susan will choose Harvard, but Billy loves Yale and has great "character" and extracurricular activities. In the end, Harvard rejects Billy and accepts Susan (decreasing acceptance rate, increasing yield). Yale rejects Susan and accepts Billy (decreasing acceptance rate, increasing yield).

The horsetrading described by @RN4407 depends on a relationship where the college can trust statements made by the high school counselor and vice-versa. This relationship is the product of repeated interactions over many years. @RN4407 writes that colleges "can reward that [high] school for good information and sanction it for bad information" [p. 173]. A high school that makes false statements about applicant intentions faces consequences. The college "might stop taking these telephone calls and ignore the information provided. They may even start accepting fewer students from the school, thinking it is less than an honest place." [CITE BOOK CHAPTER; PG NUMBER]. Because the college and the high school are mutually dependent, however, both "have an incentive to continue with a strong, honest relationship" [CITE BOOK CHAPTER; PG NUMBER].  Such horsetrading may be less pervasive than now than it was in prior decades, less common at non-elite private schools, and also public university admissions officers likely have less authority to engage in these tacit negotiations.

Nevertheless, maintaining a strong relationships is mutually important for the college and the high school.  Both the college admissions counselor and the high school guidance counselor need to tell one another "their story" and relay that story to constituents. The college admissions counselor explains why the college/university is a "special place" that high school students should want to attend. The guidance counselor explains why the high school is a special place, that even students with lower grades have outstanding extracurricular strengths that will benefit the college. [CITES FOR THIS PARAGRAPH]

Circling back to the focus of this chapter, off-campus recruiting visits are necessary for the maintenance of strong relationships that enable colleges and high schools to negotiate and send trustworthy information to one another. Without face-to-face visits, it is less likely that that a college admissions counselor will "take the call" of a guidance counselor. Simultaneously, a recruiting visit between a college and a high school can be conceived as an indicator. First, the fact that the college took time and effort to make the visit suggests that the college wants to enroll students from the high school. Second, the fact that the high school hosted the visit suggests that the high school likely views the college as a desirable destination for some of its students.  Third, the presence of the recruiting visit indicates that the high school and the college have a relationship and suggests the probability of additional interactions (e.g., phone calls).

Our project collected data about off-campus recruiting visits made in 2017 by a convenience sample of colleges universities. Our analysis sample is based on three different lists of postsecondary institutions: first, the set of public research universities (which category) as defined by the YYYY Carnegie Classification; second, private universities in the top 100 of USNWR YYYY "national universities"; and, third, private colleges in the top 50 of USNWR YYYY "national liberal arts colleges" rankings. For each of these institutions, we investigated their admissions website for pages that provided the details of upcoming off-campus recruiting visits. For institutions that posted such pages, we scraped the pages once per week throughout the 2017 calendar year. Many colleges and university only posted certain kinds of events (e.g., hotel receptions and national college fairs) but not others (e.g., day-time visits to high schools). These institutions are excluded from the analyses. Our final analysis sample consists of X public research universities, X private research universities, and X private liberal arts colleges. [MAYBE ADD TABLE/FIGURE OR APPENDIX TABLE/FIGURE ON CHARACTERISTICS OF INSTITUTIONS IN OUR SAMPLE?]

Our analyses to date have focused only on visits by public research universities [CITE]. The big-picture takeaways are as follows. Most public research universities in our sample made more out-of-state than in-state recruiting visits. The exceptions were UC-Berkeley, UC-Irvine, and North Carolina State University, three institutions with relatively generous state appropriations and the only three public universities in our sample that faced strong nonresident enrollment caps. Accross all universities, out-of-state visits systematically targeted affluent, predominantly white schools and communities, usually within populous metropoitan areas. Most public universities also made a disproportionate number of out-of-state visits to private high schools. Analyses of in-state visits also found that public high schools in affluent communities were more likely to receive a visit, but this relationship was modest compared to the income disparity between visited and not visited out-of-state public schools. Furthermore, in-state visits did not consistently indicate preferences for majority white high schools or for private high schools.

Tables/Figures X and X show simple counts and descriptive staitistics of off-campus recruiting visits by colleges and universities in our sample. TEXT BRIEFLY DESCRIBING FINDINGS FROM DESCRIPTIVE TABLES.... Table/Figure X demonstrates the disproportionate number of visits to private high schools. The goal of this chapter is to investigate off-campus recruiting visits to private high schools. Which types of private high schools receive visits from which types of colleges and universities? Additionally, to what extent were visits a college/university to a private high school more likely if the two organizations were similar to one another on key characteristics? Our analyses focus on the characteristics of academic reputation, geographic region, religious affiliation, and racial composition. WHAT TO SHOW IN TABLES/FIGURES:
<!--
- Table/figure of number of visits by visit type and in/out-of-state
- Table/figure showing disproportionate number of visits to private high schools.
-->

[PARAGRAPH ABOUT TABLE THAT DESCRIBES CHARACTERISTICS OF VISITED VS. NON-VISITED PRIVATE SCHOOLS]
Before proceeding to network analyses, Table/Figure X compares the characteristics of private high schools that received at least one recruiting visit from a college/university in our sample to those of private high schools that did not receive a visit. WHICH CHARACTERISTICS TO SHOW: geographic region; religious affiliation; academic reputation; racial composition; enrollment size.

# Network Analyses of Off-Campus Recruiting Visits

## Why Network Analysis and Network Basics

We utilize network methods rather than conventional descriptive statistics for two, related reasons. First, the theory and analytic methods of social network analysis privileges the relationship between actors. The discussion of Stevens and Khan suggests that for private school students, "it is not just the quality of the students that gets them into college but the quality of the relationship between elite high schools and colleges" [@RN4407] [pg. 175]. Second, a visit from a college/university admissions counselor to a high school can be conceived as a "network tie" connecting two actors in a social network.

We briefly define essential social network analysis concepts and their operationalization in our analyses.  A social network consists of a set of actors -- referred to as "vertices" -- and the connections -- referred to as "edges" between these actors.  Whereas "one mode" networks consist of vertices of the same "type" (e.g., in a publication network each vertex is an author), "two mode" networks consit of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an an actor-movie network consists of actors -- mode 1 -- who appear in movies -- mode 2 and an actor shares an edge with a movie if the actor appears in the movie.  The social network analyzed in this chapter is a two-mode network, where vertices consists of colleges/universities (mode 1) and high schools (mode 2) and an "edge" is defined as a visit from a college/university to a private high school.

Our analyses draw from analyses of "corporate board-director" networks because these are the most commonly analyzed two-mode networks in the social networks literature. Board-director networks consist of directors (mode 1) who serve on organizational boards (mode 2). @RN4663 states that "Overlapping groups such as boards of directors form a two-mode membership network in which one can conceive of directors as nodes [vertices] connected by a tie of common board membership, or boards as nodes connected by a tie of one or more shared directors" [pg. xx].  Similarly, our college-school network can be conceive of colleges as vertices that are connected to high school verticies by a tie of an off-campus recruiting visit.  In practice, two-mode networks are ofen analyzed as one-mode networks. In the corporate board interlock literature a one-mode director-by-director matrix is created by identifying whether two directors serve on one ore more board (e.g., two directors share a tie if both serve on the board of Google) and a one-mode company-by-company matrix is created by identifying whether two boards have one or more directors who serve on both boards (e.g., xxxxxx). Similarly, we create a one-mode college network in which two colleges share a connection if both colleges visit at least one high school in common and a one-mode high school network in which two high schools share an edge if they are visited by at least one college/university in common.

A two-mode board-director network can also be conceived as two one-mode networks.

In the social network analysis literature, the corporate board-director network is most commonly analyzed two-mode network [CITE]. These networks consist of directors (mode 1) who serve on organizational boards (mode 2).

The most commonly analyzed two-mode network is the "corporate board interlock."

For example, in a two-mode actor-movie network consists of two modes of vertices, actors and movies.
